Optimized Multi-Frequency Spectra for Numerical Radiative Transfer

### 2011-10-26 doc/paper/ms.tex
Got rid of rt1d.py hopefully (case-folding collision).

%%% 
%% Multi-Frequency vs. Continuous
%%%
\section{ASSESSING THE CONSEQUENCES OF DISCRETE RADIATION FIELDS}
Before computing the optimal multi-frequency spectra for SEDs most commonly used in numerical simulations (\S3), we will first demonstrate the difference in results brought about by varying degrees of frequency resolution.  To do this, we use \texttt{rt1d}, a 1D radiative transfer code developed for this work that is described and validated in the appendices.

@@

\section{METHODS}
Upon careful inspection of the coupled rate equations which govern the balance
of ionizations, recombinations, heating, and cooling, one realizes that the
integral values, i.e. equations \ref{eq:HIIRateEquation},
\ref{eq:HeIIRateEquation}, \ref{eq:HeIIIRateEquation}, and
\ref{eq:TemperatureEvolution} can be tabulated as a function of column density
(i.e. optical depth), allowing for a significant speed-up in run time. The key
to our subsequent analysis was inspired by this same convenient feature: if
$\Gamma_i$ and $\Heat_i$ can be tabulated for a given source
\textit{a-priori}, it means that \textit{precision determination of the
ionization state and temperature of the gas surrounding that source relies
solely on accurate numerical computation of the photoionization and heating
rate integrals in Equations \ref{eq:PhotoionizationRate} and
\ref{eq:HeatingRate}}. So, for a given source, we can compute the values of
these integrals numerically for the continuous spectrum, and then as discrete
sums given $n_{\nu}$ frequency bins with
\begin{equation}
    I(E) = \sum_{j=1}^{n_{\nu}} I(E_j) .
\end{equation}

% Discretization Artifacts
\subsection{Discretization Artifacts}

@@

\section{RESULTS}

@@

Each of these equations represents the balance of ionizations and recombinations for ion $i = \HII, \HeII, \HeIII$, where $\Gamma_i$ denotes the photo-ionization rate coefficient, $\alpha_i$ ($\xi_i$) the case-A (dielectric) recombination rate coefficients, and $\beta_i$ the collisional ionization rate coefficients.  In this work, we use the formulae in Appendix B of \citet{Fukugita1994} to compute the values of $\alpha$, $\beta$, and $\xi$.    We absorb ionizations due to to energetic photo-electrons into the  photoionization rate coefficients, $\Gamma_i$, which are given by

@@

where $E_i$, $\sigma_i$, $n_i$, and $f_i$ are the ionization threshold energy, bound-free absorption cross section, number density, and fraction of photo-electron energy that goes into ionization of species $i$, respectively, and the sum over $j = \HI, \HeII, \HeII$ term represents ionizations due to fast secondary electrons from photoionizations of species $j$, which has number density $n_j$.  $\spec$ is the spectral intensity, which we write as

@@

and $\zeta_i$, $\eta_i$, and $\psi_i$ are the collisional ionization, recombination, and collisional excitation cooling coefficients, respectively.  Additional parameters include $\nb$, the total number density of baryons (check on this), the mean molecular weight $\mu$, Boltzmann's constant $\kB$, the kinetic temperature of the gas $T$, and $\fheat$, the fraction of a photon's original energy deposited as heat (see \citet{Shull1985}, \citet{Furlanetto2010}).  For simplicity, we neglect cosmological effects, including Hubble cooling due to cosmological expansion, Compton heating/cooling off CMB photons, and photo-ionization by CMB photons. Currently neglecting cooling by free-free emission too.

@@

It is now apparent that the values of $\Gamma_i$ and $\Heat_i$ can be tabulated as a function of column density for a given source.  This greatly improves the speed of 1D radiative transfer calculations, like those of \cite{Thomas2008}, since the integrals in these expressions (Equations \ref{eq:PhotoionizationRate} and \ref{eq:HeatingRate}) no longer need to be computed numerically on-the-fly.  However, the lookup tables will only be 1-dimensional for the case of a hydrogen-only medium.

@@

At this juncture it is worth mentioning the numerical artifacts that can result due to poor temporal or spatial resolution.

% I-front evolution, RT06 #1
\begin{figure*}[htbp]
\centering
\includegraphics[scale=0.8]{figures/RT06_1_IfrontEvolution.eps}
\caption{\textit{Top:} Comparison of the numerical (dashed) and analytic (solid) solutions for the position of an expanding ionization front as a function of time in a hydrogen-only, isothermal medium (RT06 problem \#1).  The numerical solution displayed here is from the highest resolution simulation (6400 grid cells, i.e. $\Delta x = L_{\mathrm{box}}/6400$). \textit{Bottom:} Ratio of the calculated and analytic solutions as a function of time and grid resolution.  Simulations with > 200 grid cells converge on the highest resolution solution to an accuracy of $\lesssim 5 \%$.}
\label{fig:RT_Test1_IfrontEvolution}
\end{figure*}

To demonstrate the functionality of the code, we repeat tests 1 and 2 from the Radiative Transfer Comparison Project (\cite{Iliev2006}; hereafter referred to as RT06).  Test 1 is the expansion of an HII region in a hydrogen-only, isothermal medium surrounding a monochromatic source of 13.6 eV photons. We adopt the parameters used in \citet{Wise2011}: constant temperature $T = 10^4 \ \mathrm{K}$, uniform hydrogen number density $n_H = 10^{-3} \ \mathrm{cm^{-3}}$, ionized fraction $x = 1.2 \times 10^-3$, in a box $L_{\mathrm{box}} = 6.6 \ \mathrm{kpc}$ in size, and with photon luminosity $\dot{Q} = 5\times10^{48} \ \mathrm{s^{-1}}$.  The classical analytic solution for the radius of an ionization front is 
\begin{equation}
    r_{\mathrm{IF}}(t) = r_s (1 - e^{-t/t_{\mathrm{rec}}})^{1/3}
\end{equation}    
where $r_s$ is the Str\"{o}mgren radius, 
\begin{equation}
    r_s = \left(\frac{3 \dot{Q}}{4\pi \alpha_{\mathrm{HII}}(T) n_{\mathrm{HI}}^2}\right)^{1/3} ,
\end{equation}
and the recombination time, \trec, is defined as
\begin{equation}
    t_{\mathrm{rec}} = \frac{1}{\alpha_{\mathrm{HII}}(T) n_{\mathrm{HI}}} .
\end{equation}
This solution is approximate even in isothermal media, given that it
assumes a constant neutral hydrogen density, $\nHI$. More accurate analytic
solutions exist (citations), and predict a departure from the classical
solution at $t/t_{\mathrm{rec}} \simeq 1$, which grows to a $\sim 5 \%$
difference by $t/t_{\mathrm{rec}} \simeq 4$. Our numerical solution (see
Figure \ref{fig:RT_Test1_IfrontEvolution}) captures this behavior very well.  In Figure \ref{fig:RT_Test1_RadialProfiles}, we show radial profiles of the ionized and neutral fractions at three stages of the I-front expansion, again in very good agreement with the calculations in \citet{Iliev2006}.

>>>>

%%% 
%% Multi-Frequency vs. Continuous
%%%
\section{ASSESSING THE CONSEQUENCES OF DISCRETE RADIATION FIELDS} \label{sec:Consequences}
Before computing the optimal multi-frequency spectra for SEDs most commonly used in numerical simulations (\S3), we will first demonstrate the difference in results brought about by varying degrees of frequency resolution.  To do this, we use \texttt{rt1d}, a 1D radiative transfer code developed for this work that is described and validated in the appendices.

@@

\section{METHODS} \label{sec:Methods}
Upon careful inspection of the coupled rate equations which govern the balance
of ionizations, recombinations, heating, and cooling, one realizes that the
integral values, i.e. equations \ref{eq:HIIRateEquation},
\ref{eq:HeIIRateEquation}, \ref{eq:HeIIIRateEquation}, and
\ref{eq:TemperatureEvolution} can be tabulated as a function of column density
(i.e. optical depth), allowing for a significant speed-up in run time. The key
to our subsequent analysis was inspired by this same convenient feature: if
$\Gamma_i$ and $\Heat_i$ can be tabulated for a given source
\textit{a-priori}, it means that \textit{precision determination of the
ionization state and temperature of the gas surrounding that source relies
solely on accurate numerical computation of the photoionization and heating
rate integrals in Equations \ref{eq:PhotoionizationRate} and
\ref{eq:HeatingRate}}. So, for a given source, we can compute the values of
these integrals numerically for the continuous spectrum, and then as discrete
sums given $n_{\nu}$ frequency bins with
\begin{equation}
    I(E) = \sum_{j=1}^{n_{\nu}} I(E_j) .
\end{equation}

@@

\section{RESULTS} \label{sec:Results}

@@

Each of these equations represents the balance of ionizations and
recombinations for ions $\HII$, $\HeII$, $\HeIII$. We associate the index $i$
with neutral species, $i = \HI, \HeI, \HeII$, and define $\Gamma_i$ as the
photo-ionization rate coefficient, $\alpha_i$ ($\xi_i$) the case-A
(dielectric) recombination rate coefficients, $\beta_i$ the collisional
ionization rate coefficients, and $\nel$ the number density of electrons,
    \nel = \nHII + \nHeII + 2\nHeIII .
In this work, we use the formulae in Appendix B of \citet{Fukugita1994} to compute the values of $\alpha$, $\beta$, and $\xi$.    We absorb ionizations due to to energetic photo-electrons into the  photoionization rate coefficients, $\Gamma_i$, which are given by

@@

where $E_i$, $\sigma_i$, $n_i$, and $f_i$ are the ionization threshold energy, bound-free absorption cross section, number density, and fraction of photo-electron energy that goes into ionization of species $i$, respectively. The term summing over $j$, where $j = \HI, \HeI, \HeII$, represents ionizations of species $i$ due to fast secondary electrons from photoionizations of species $j$, which has number density $n_j$.  $\spec$ is the spectral intensity, which we write as

@@

and $\zeta_i$, $\eta_i$, and $\psi_i$ are the collisional ionization, recombination, and collisional excitation cooling coefficients, respectively. The total number density of baryons is given by $\nb$, the mean molecular weight, $\mu$, Boltzmann's constant, $\kB$, the kinetic temperature of the gas, $T$, and the fraction of a photon's original energy deposited as heat, $\fheat$ (see \citet{Shull1985}, \citet{Furlanetto2010}).  For simplicity, our current treatment neglects cooling via free-free emission, as well as cosmological effects, including Hubble cooling due to cosmological expansion, Compton heating/cooling off cosmic microwave background (CMB) photons, and photo-ionization by Wien-tail CMB photons.

@@

It is now apparent that the values of $\Gamma_i$ and $\Heat_i$ can be tabulated as a function of column density for a given source.  This greatly improves the speed of 1D radiative transfer calculations, like those of \cite{Thomas2008}, since the integrals in these expressions (Equations \ref{eq:PhotoionizationRate} and \ref{eq:HeatingRate}) no longer need to be computed numerically on-the-fly.  It also provides the basis for our frequency resolution optimization strategy, as presented in \S\ref{sec:Methods}. Note, however, that the dimensionality of these lookup tables is equal to the number of absorbing species, so the tables for simulations including only hydrogen are 1D, and those including helium are 3D.

@@

%At this juncture it is worth mentioning the numerical artifacts that can result due to poor temporal or spatial resolution.

% I-front evolution, RT06 #1
\begin{figure*}[htbp]
\centering
\includegraphics[scale=0.8]{figures/RT06_1_IfrontEvolution.eps}
\caption{\textit{Top:} Comparison of the numerical (dashed) and analytic (solid) solutions for the position of an expanding ionization front as a function of time in a hydrogen-only, isothermal medium (RT06 problem \#1).  The numerical solution displayed here is from the highest resolution simulation (6400 grid cells, i.e. $\Delta x = L_{\mathrm{box}}/6400$). \textit{Bottom:} Ratio of the calculated and analytic solutions as a function of time and grid resolution.  Simulations with $>$ 200 grid cells converge on the highest resolution solution to an accuracy of $\lesssim 5 \%$.}
\label{fig:RT_Test1_IfrontEvolution}
\end{figure*}

To demonstrate the functionality of the code, we repeat tests 1 and 2 from the Radiative Transfer Comparison Project (\cite{Iliev2006}; hereafter referred to as RT06).  Test 1 is the expansion of an HII region in a hydrogen-only, isothermal medium surrounding a monochromatic source of 13.6 eV photons. We adopt the parameters used in \citet{Wise2011}: constant temperature $T = 10^4 \ \mathrm{K}$, uniform hydrogen number density $n_H = 10^{-3} \ \mathrm{cm^{-3}}$, ionized fraction $x = 1.2 \times 10^-3$, in a box $L_{\mathrm{box}} = 6.6 \ \mathrm{kpc}$ in size, and with photon luminosity $\dot{Q} = 5\times10^{48} \ \mathrm{s^{-1}}$.  The classical analytic solution for the radius of an ionization front is 
\begin{equation}
    r_{\mathrm{IF}}(t) = r_s (1 - e^{-t/t_{\mathrm{rec}}})^{1/3}
\end{equation}    
where $r_s$ is the Str\"{o}mgren radius, 
\begin{equation}
    r_s = \left(\frac{3 \dot{Q}}{4\pi \alpha_{\mathrm{HII}}(T) n_{\mathrm{HI}}^2}\right)^{1/3} ,
\end{equation}
and the recombination time, \trec, is defined as
\begin{equation}
    t_{\mathrm{rec}} = \frac{1}{\alpha_{\mathrm{HII}}(T) n_{\mathrm{HI}}} .
\end{equation}
This solution is approximate even in isothermal media, given that it
assumes a constant neutral hydrogen density, $\nHI$. More accurate analytic
solutions exist (citations), and predict a departure from the classical
solution at $t/t_{\mathrm{rec}} \simeq 1$, which grows to a $\sim 5 \%$
difference by $t/t_{\mathrm{rec}} \simeq 4$. Our numerical solution (see
Figure \ref{fig:RT_Test1_IfrontEvolution}) captures this behavior very well.  In Figure \ref{fig:RT_Test1_RadialProfiles}, we show radial profiles of the ionized and neutral fractions at three stages of the I-front expansion, which are again in very good agreement with the calculations presented in \citet{Iliev2006}.
